---
title: "Water Temperature and Juvenile Chinook Salmon Fork Length in The San Francisco Estuary"
description: |
  Investigating whether the mean fork length of juvenile Chinook salmon in the San Francisco Estuary differs at sample sites with colder water temperatures compared to warmer water temperatures.
author:
  - name: Charles Hendrickson
date: 12-02-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F, fig.pos='H')
# Load all the packages needed here
library(tidyverse)
library(readr)
library(gt)
library(tufte)
library(feasts)
library(tsibble)
library(xtable)
library(knitr)
```

# Research Question

The aim of this project is to investigate whether the mean fork length of juvenile Chinook salmon in the San Francisco Estuary differs at sample sites with colder water temperatures compared to warmer water temperatures. 

# Introduction 

Chinook salmon Oncorhynchus tshawytscha are anadromous fish, which means they are born in freshwater rivers and migrate as juveniles to saltwater to feed, grow, and mature before returning to freshwater to spawn (NOAA Fisheries). They are vulnerable to many anthropogenic stressors, such as habitat degradation, blocked access to spawning grounds, and loss of genetic diversity from hatcheries. A particularly concerning stressor for Chinook salmon are changes in freshwater thermal regimes due to simplification of the physical structure of aquatic systems, reduced shading from the removal of riparian vegetation, and altered flow regimes from climate change and water diversion. (Richter & Kolmes, 2005). In a study that investigated the effects of three temperature regimes on juvenile Chinook salmon growth, it was found that juveniles reared at an elevated temperature range of 21–24°C experienced significantly decreased growth rates compared with juveniles reared at 13–16°C (Marine & Cech Jr, 2004). Furthermore, studies have found that juvenile Chinook salmon length is positively correlated with the survival and survival was markedly low among the smallest length class (that is, 80 mm to 89 mm) of subyearling Chinook salmon (Brown, Oldenburg & Seaburg et al., 2013). As California’s Chinook salmon populations become increasingly threatened by changes in freshwater thermal regimes and with increasing evidence suggesting that juvenile Chinook Salmon survival could be negatively impacted by decreased fork length, it is necessary to understand how the size of out-migrating juvenile Chinook salmon in the San Francisco Estuary may be impacted by the water temperature of their native rivers. 

While previous studies have used controlled laboratory experiments to test the effects of water temperature on Chinook salmon growth, I will use a long-term fish monitoring dataset in my analysis to quantify the impact of water temperature on juvenile Chinook salmon fork length.


# Data Description and Collection

I downloaded Interagency Ecological Program (IEP) survey data from the Environmental Data Initiative (EDI) Data Portal. 

* [Interagency Ecological Program survey data](https://portal.edirepository.org/nis/mapbrowse?packageid=edi.244.8) 

The IEP survey data contains over four decades (1976-2021) of juvenile fish monitoring data from the San Francisco Estuary, collected by the The United States Fish and Wildlife Service Delta Juvenile Fish Monitoring Program (DJFMP). As part of the IEP that manages the San Francisco Estuary, the DJFMP has tracked the relative abundance and distribution of naturally and hatchery produced juvenile Chinook Salmon of all races as they out-migrate through the Sacramento-San Joaquin Delta using a combination of surface trawls and beach seines. Since 2000, three trawl sites and 58 beach seine sites have been sampled weekly or biweekly within the San Francisco Estuary and lower Sacramento and San Joaquin Rivers. 
The DJFMP data set contains columns of the sample site location, date, time, sample method, environmental conditions, water temp, species, fish count, and fork length. Each row represents a sample of one or multiple fish (must have the same fork length) at a distinct site and time. The strengths of this data set are that the sample size is very large and has ample metadata. The limitations of this data set are that fork length samples have not been taken continuously since the beginning so there are gaps in the data. Also the data set is not already in tidy format because the ‘Count’ column contains the number of fish sampled with the same measured fork length. The full metadata can be found [here](https://portal.edirepository.org/nis/metadataviewer?packageid=edi.244.8).
  
The variables that I am interested in for my analysis are fork length (mm) and water temperature (degrees Celsius). 

- **Fork Length**

Fork length is the length of sampled fish measured from the point of the mouth to the fork of the caudal fin. Fish were measured to the nearest 1 mm before being released. Generally, all fish in each sample that were ≥ 25 mm fork length were identified by species. 
 


```{r}
# URL of data set: https://portal.edirepository.org/nis/metadataviewer?packageid=edi.244.8

# Import data and convert it to a data frame
DJFMP_df <- as.data.frame(
  read.csv(file = "data/1976-2021_DJFMP_beach_seine_fish_and_water_quality_data.csv"))

```


```{r}
# Filter data set for just Chinook salmon (CHN) observations 
DJFMP_CHN_df <- DJFMP_df %>% 
  filter(OrganismCode == 'CHN') %>% 
  select(Location, SampleDate, ForkLength, WaterTemp, MethodCode, Count) %>% 
  mutate(SampleDate = as.Date(SampleDate))

# Drop NA's from WaterTemp column
DJFMP_CHN_df <- DJFMP_CHN_df[!is.na(DJFMP_CHN_df$WaterTemp),]

```


```{r}
# Check for outliers
large_outliers <- DJFMP_CHN_df %>% 
  filter(ForkLength > 250)

small_outliers <- DJFMP_CHN_df %>% 
  filter(ForkLength <= 0)

# Remove these data points because juvenile Chinook salmon cannot have a fork length of zero mm and we only want data on juvenile Chinook salmon that are below 250 mm in length, not adults. 
DJFMP_CHN_df <- DJFMP_CHN_df %>% 
  filter(ForkLength != 0 & ForkLength < 250)
```


```{r, fig.width=6, fig.height=3, fig.cap="Fork length of juvenile Chinook salmon in San Francisco Estuary from 1976 to 2021."}
# Plot juvenile Chinook salmon fork length over time
ggplot(data = DJFMP_CHN_df, aes(x = SampleDate, y = ForkLength)) +
  geom_point() +
  labs(title = "Fork Length of Juvenile Chinook Salmon 
       in San Francisco Estuary (1976-2021)",
       x = "Date",
       y = "Fork Length (mm)") +
  theme_classic()
```
 
- **Water Temperature**

The YSI PRO 2030 Meter was used to measure water temperature before or after each fish sample. Water temperature was recorded to the nearest 0.1˚C. 
 
```{r, fig.width=6, fig.height=3, fig.cap="Water Temperature at Sample Sites in San Francisco Estuary from 1976 to 2021."}
# Plot water temperature over time
ggplot(data = DJFMP_CHN_df, aes(x = SampleDate, y = WaterTemp)) +
  geom_point() +
  labs(title = "Water Temperature Across Sample Sites in 
       San Francisco Estuary (1976-2021)",
       x = "Date",
       y = "Water Temperature (degrees Celsius)") +
  theme_classic()
```

# Analysis

I will conduct a hypothesis test to determine if there is a difference in the mean juvenile Chinook salmon fork length (mm) across sites with warmer or colder than average water temperatures in the San Francisco Estuary. Specifically, I will use a linear regression to see if water temperature affects fork length. 
First, I will define my null and alternative hypotheses. Second, I will collect, tidy, filter, and visualize my data. Third, I will calculate the average water temperature across all samples and then subset data into warmer than average and colder than average sites. Then, I will compute a point estimate of the difference between the mean fork length of juvenile Chinook salmon that are reared in warmer than average sites versus colder than average sites. Fourth, I will conduct a t-test to measure the variability of the sample statistic. Fifth, I will quantify the probability that the sample statistic differs from the null hypothesis by running a linear regression of water temperature on fork length. Finally, based on my p-value, I will reject or fail to reject my null hypothesis. 

# Limitations

A limitation of my analysis is that water temperature is the only independent variable included in the model, which means this analysis is vulnerable to omitted variables bias. It is very likely that there are many different abiotic and biotic factors in addition to water temperature that influence juvenile Chinook salmon fork length. 

# Hypotheses

**Null Hypothesis**
$$H_{0}: \mu_{ForkLengthCold} - \mu_{ForkLengthWarm} = 0$$


**Alternative hypothesis**
$$H_{A}: \mu_{ForkLengthCold} - \mu_{ForkLengthWarm} \neq 0$$


# Visualize data  

Scatter plot showing changes in juvenile Chinook salmon fork length ($y$-axis) as it relates to water temperature ($x$-axis).

```{r}
forklength_vs_watertemp <- ggplot(data = DJFMP_CHN_df, 
                                  aes(x = WaterTemp, 
                                      y = ForkLength)) +
  geom_point(size = 1) +
  labs(title = "Relationship Between Juvenile Chinook Salmon Fork Length and Water Temp",
       x = "Water Temperature (degrees C) ",
       y = "Fork Length (mm)") +
  theme_bw()

forklength_vs_watertemp
```

Density plot of juvenile Chinook salmon fork length ($y$-axis) as it relates to water temperature ($x$-axis).

```{r}
# Determine where data density is relatively high versus low
forklength_vs_watertemp_alpha <- ggplot(data = DJFMP_CHN_df, 
                                        aes(x = WaterTemp, 
                                            y = ForkLength)) +
  geom_point(alpha=0.1, size=1) +
  labs(title = "Relationship Between Juvenile Chinook Salmon Fork Length and Water Temp",
       x = "Water Temperature (degrees C) ",
       y = "Fork Length (mm)") +
  theme_bw() 

forklength_vs_watertemp_alpha
```

Density plot with `geom_bin2d()` of juvenile Chinook salmon fork length ($y$-axis) as it relates to water temperature ($x$-axis).

```{r}
forklength_vs_watertemp_density <- ggplot(data = DJFMP_CHN_df, 
                                  aes(x = WaterTemp, 
                                      y = ForkLength)) +
  geom_bin2d() +
  labs(title = "Relationship Between Juvenile Chinook Salmon Fork Length and Water Temp",
       x = "Water Temperature (degrees C) ",
       y = "Fork Length (mm)") +
  theme_bw()

forklength_vs_watertemp_density
```


# Compute Point Estimate 

```{r}
# Mean water temp of all sites 
mean_water_temp_overall <- mean(DJFMP_CHN_df$WaterTemp)

# Get the count of warm water sites vs cold water sites
warm_count <- as.numeric(
  count(DJFMP_CHN_df %>% filter(WaterTemp > mean_water_temp_overall)))

cold_count <- as.numeric(
  count(DJFMP_CHN_df %>% filter(WaterTemp < mean_water_temp_overall)))


# Mean fork length of warm water sites and cold water sites
warm_sites <- DJFMP_CHN_df %>%
  filter(WaterTemp > mean_water_temp_overall) %>%
  summarise(mean = mean(ForkLength))
  
cold_sites <- DJFMP_CHN_df %>% 
  filter(WaterTemp < mean_water_temp_overall) %>%
  summarise(mean = mean(ForkLength))

# Difference in mean fork lengths for cold water sites vs. warm water sites 
point_est = as.numeric(warm_sites - cold_sites)
print(point_est)

```

The point estimate for the difference in mean fork lengths for cold water sites versus warm water sites  is 7.342742 mm. However, this value could just be a difference in the samples due to random variability. Thus, we must calculate a measure of variability, not just a measure of the the difference in means.


# Model the variability of the sample statistic

Conduct t-test

```{r}
t.test(DJFMP_CHN_df$ForkLength[DJFMP_CHN_df$WaterTemp > mean_water_temp_overall], 
       DJFMP_CHN_df$ForkLength[DJFMP_CHN_df$WaterTemp < mean_water_temp_overall])
```

The **test statistic** (z-score) tells us that the observed difference in the mean fork length for cold water and warm water sites is 74.038 standard deviations above the null hypothesis of zero difference.

The **95% confidence interval** tells us that there is a 95% probability that the range [7.148360, 7.537123] is likely to contain the true population parameter, which is the difference in juvenile Chinook salmon mean fork lengths across sites with colder water temperatures versus sites with warmer water temperatures.


# Quantify the probability that the sample statistic differs from the null

- Use the `lm()` command to estimate the following simple linear regression:
$$ \text{fork length}_i = \beta_0 + \beta_1 \text{water temperature}_i + \varepsilon_i $$
Run a linear regression of y on x

```{r}
summary(lm(ForkLength ~ WaterTemp, data = DJFMP_CHN_df))
```

Based on our low p-value, we can reject the null hypothesis that there is no effect of water temperature on fork length. 

**Intercept:** On average, the predicted fork length of a juvenile Chinook salmon for a zero degree Celsius change in water temperature is 30.3742 (mm).  

**Coefficient on water temperature (WaterTemp):** On average, the predicted fork length of a juvenile Chinook salmon increases by 1.6276 (mm) for every one degree Celsius increase in water temperature. 

**R-squared:** Our coefficient of determination ($R^2$) tells us that 7.75% of the variation in juvenile Chinook salmon fork lengths is explained by water temperature at the site. 


# Visualize Regression Line

```{r}
forklength_vs_watertemp <- ggplot(data = DJFMP_CHN_df, 
                                  aes(x = WaterTemp, 
                                      y = ForkLength)) +
  geom_point() +
  geom_smooth(method = 'lm', formula= y~x, se=FALSE, size=1) +
  labs(title = "Relationship Between Juvenile Chinook Salmon Fork Length and Water Temp",
       x = "Water Temperature (degrees C) ",
       y = "Fork Length (mm)") +
  theme_bw()

forklength_vs_watertemp
```


# Evaluate whether we reject or fail to reject the null hypothesis 

Since $p-value = 2.2e-16 < 0.05$ we reject the null that there is no difference in the mean juvenile Chinook salmon fork length at sites with colder water temperatures versus sites with warmer water temperatures. We can say there is a statistically significant difference (at the 5% significance level) in the juvenile Chinook salmon mean fork length across sites with colder water temperatures versus sites with warmer water temperatures.

